---
title: "Baseball Analysis"
author: "Injoong Kim and Stockton Aubrey"
format: 
  html: 
    embed-resources: true
editor: visual
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyverse)
library(Lahman)
library(plotly)
library(corrplot)
library(car)
```

## PROMPT

Baseball is a great sport, and we are big fans of the game, and with the aid of Lahman's Database, we are selecting 50 of our favorite baseball players and determining if height and weight affects how many homeruns are hit over the last 10 years of their career and their performance on fielding. Additionally, we seek to ascertain whether a player's handedness significantly impacts their batting average across the same duration.

The datasets used in this analysis include the People, Batting, and Fielding datasets from Lahman's Database, encompassing crucial variables such as playerID, nameFirst, nameLast, weight, height, bats, debut, finalGame from the People dataset, playerID, yearID, PO(Putouts), A(Assist), E(Error) from the Fielding dataset,and the playerID, AB(At Bats), H(Hits), and HR(Homerun) variables from the Batting dataset. To ensure a comprehensive and accurate analysis, the 10 seasons under consideration will be the 10 seasons leading up to each player's final game season, since rookie seasons often lack sufficient data to draw meaningful conclusions.

Through this rigorous statistical investigation, we aim to uncover potential relationships between physical attributes and players' batting and fielding performance, contributing to the ongoing discourse on the multifaceted dynamics shaping success in professional baseball.

## IMPORT

```{r}
#Read the 'People' Dataset and select the variables we need from the 'People' Dataset 
people <- subset(People, select = c(playerID, nameFirst, nameLast, weight, height, bats, debut, finalGame))
tail(people, n=10)

#Read the 'Batting' Dataset and select the variables we need from the 'Batting' Dataset 
batting <- subset(Batting, select = c(playerID, yearID, AB, H, HR))
tail(batting, n=10)

#Read the 'Fielding' Dataset and select the variables we need from the 'Fielding' Dataset 
fielding <- subset(Fielding, select = c(playerID, yearID, PO, A, E))
tail(fielding, n=10)
```

## TIDY

```{r}
#Select the 50 players who are our favorite players from the 'People' Dataset 
players <- people |>
  filter((nameFirst == "Mike" & nameLast == "Trout") |
           (nameFirst == "Bryce" & nameLast == "Harper") |
           (nameFirst == "Matt" & nameLast == "Kemp") |
           (nameFirst == "Ichiro" & nameLast == "Suzuki") |
           (nameFirst == "Derek" & nameLast == "Jeter") |
           (nameFirst == "Babe" & nameLast == "Ruth") |
           (nameFirst == "Ted" & nameLast == "Williams") |
           (nameFirst == "Barry" & nameLast == "Bonds") |
           (nameFirst == "Manny" & nameLast == "Machado") |
           (nameFirst == "Fernando" & nameLast == "Tatis") |
           (nameFirst == "George" & nameLast == "Brett") |
           (nameFirst == "Albert" & nameLast == "Pujols") |
           (nameFirst == "Ryan" & nameLast == "Howard") |
           (nameFirst == "Buster" & nameLast == "Posey") |
           (nameFirst == "Pete" & nameLast == "Rose" & debut == "4/8/1963") |
           (nameFirst == "David" & nameLast == "Ortiz") |
           (nameFirst == "Mike" & nameLast == "Schmidt") |
           (nameFirst == "Paul" & nameLast == "Goldschmidt") |
           (nameFirst == "Todd" & nameLast == "Helton") |
           (nameFirst == "Matt" & nameLast == "Holliday") |
           (nameFirst == "Alex" & nameLast == "Rodriguez") |
           (nameFirst == "Sammy" & nameLast == "Sosa") |
           (nameFirst == "Mark" & nameLast == "McGwire") |
           (nameFirst == "Mickey" & nameLast == "Mantle") |
           (nameFirst == "Carlos" & nameLast == "Beltran") |
           (nameFirst == "Carl" & nameLast == "Yastrzemski") |
           (nameFirst == "Adrian" & nameLast == "Beltre") |
           (nameFirst == "Hank" & nameLast == "Aaron") |
           (nameFirst == "Willie" & nameLast == "Mays") |
           (nameFirst == "Prince" & nameLast == "Fielder") |
           (nameFirst == "Lou" & nameLast == "Gehrig") |
           (nameFirst == "Manny" & nameLast == "Ramirez") |
           (nameFirst == "Wade" & nameLast == "Boggs") |
           (nameFirst == "Vladimir" & nameLast == "Guerrero" & playerID == "guerrvl01")|
           (nameFirst == "Miguel" & nameLast == "Cabrera") |
           (nameFirst == "Joe" & nameLast == "Mauer")|
           (nameFirst == "Yadier" & nameLast == "Molina")|
           (nameFirst == "Chipper" & nameLast == "Jones")|
           (nameFirst == "Joey" & nameLast == "Votto")|
           (nameFirst == "Freddie" & nameLast == "Freeman")|
           (nameFirst == "Roberto" & nameLast == "Clemente")|
           (nameFirst == "Jose" & nameLast == "Altuve")|
           (nameFirst == "Harmon" & nameLast == "Killebrew")|
           (nameFirst == "Ty" & nameLast == "Cobb")|
           (nameFirst == "Jackie" & nameLast == "Robinson")|
           (nameFirst == "Tony" & nameLast == "Gwynn" & debut == "7/19/1982")|
           (nameFirst == "Cal" & nameLast == "Ripken" & debut == "8/10/1981")|
           (nameFirst == "Shin-Soo" & nameLast == "Choo")|
           (nameFirst == "Adrian" & nameLast == "Gonzalez")|
           (nameFirst == "Carl" & nameLast == "Crawford")|
           (nameFirst == "Mookie" & nameLast == "Betts")|
           (nameFirst == "Aramis" & nameLast == "Ramirez")
  )
players <- with(players,  players[order(nameFirst) , ])
players
```

## TRANSFORM

```{r}
# Join players and batting dataset for and filter the last 10 seasons of the players
battingData <- left_join(players, batting, by = join_by(playerID))|>
  dplyr::select(yearID, playerID, nameFirst, nameLast, weight, height, bats, AB, H, HR)|>
  group_by(playerID)|>
  filter(yearID >= max(yearID)-9)
battingData

#Get the total homeruns of each player
bodyHR <- battingData|>
  summarize(sumHR = sum(HR))|>
  left_join(people) |>
  dplyr::select(playerID, nameFirst, nameLast, weight, height, sumHR)
bodyHR

#Get the batting average of each player
handBatAvg <- battingData|>
  summarize(BatAvg =  round(sum(H)/sum(AB), 3))|>
  left_join(people) |>
  dplyr::select(playerID, nameFirst, nameLast, bats, BatAvg)
handBatAvg


# Join players and fielding dataset for and filter the last 10 seasons of the players
fieldingData<- left_join(players, fielding, by = join_by(playerID))|>
  dplyr::select(yearID, playerID, nameFirst, nameLast, weight, height, PO, A, E)|>
  group_by(playerID)|>
  filter(yearID >= max(yearID)-9)
fieldingData

#Get the fielding percentage of each player
fieldingPercent <- fieldingData|>
  summarize(FieldingPercentage =  round((sum(PO)+sum(A))/(sum(PO)+sum(A)+sum(E)), 3))|>
  left_join(people) |>
  dplyr::select(playerID, nameFirst, nameLast, weight, height, FieldingPercentage)
fieldingPercent
```

## VISUALIZE / MODEL

#### Impact of Weight and Height on Homeruns

```{r}
#Between Weight and Homeruns
bodyHRweight <- ggplot(bodyHR, aes(x=sumHR, y=weight))+
  geom_point()+
  labs(title = "Comparison of Homeruns and Players' Weight",
       x ="Number of Home Runs",
       y="Players' Weight")+
  theme_minimal(
  )

bodyHRweight

#Between Height and Homeruns
bodyHRheight <- ggplot(bodyHR, aes(x=sumHR, y=height))+
  geom_point()+
  labs(title = "Comparison of Homeruns and Players' Height",
       x ="Number of Home Runs",
       y="Players' Height")+
  theme_minimal(
  )

bodyHRheight

#Between Weight, Height, and Homeruns
bodyHRboth <- ggplot(bodyHR, aes(x=height, y=weight, color = sumHR, size = sumHR))+
  geom_point(alpha=0.7)+
  labs(title = "Comparison of Homeruns and Players' Height and Weight",
       x ="Players' Height",
       y ="Players' Weight")+
  theme_minimal(
  )
bodyHRboth

#3D version of bodyHRboth
bodyHRboth3D <- plot_ly(x=bodyHR$height,y=bodyHR$weight,z=bodyHR$sumHR) |>
  layout(scene = list(
    xaxis = list(title = "Height"),
    yaxis = list(title = "Weight"),
    zaxis = list(title = "Total Number of Homeruns")))

bodyHRboth3D
```

#### Batting Average by Handedness

```{r}
#Batting Average by Handedness
handedAvg <- ggplot(handBatAvg, aes(y=BatAvg, x= bats, fill=bats))+
  geom_boxplot()+
  labs(title = "Batting Average by Players' Handedness",
       x =NULL,
       y="Players' Career Batting Average")+
  theme_minimal(
  )

handedAvg

meanR <- round(mean(handBatAvg$BatAvg[handBatAvg$bats == 'R'], na.rm = TRUE),3)
meanR

meanL <- round(mean(handBatAvg$BatAvg[handBatAvg$bats == 'L'], na.rm = TRUE),3)
meanL

meanB <- round(mean(handBatAvg$BatAvg[handBatAvg$bats == 'B'], na.rm = TRUE),3)
meanB

#Number of right-handed, left-handed, and both-handed hitters in the dataset
numbats <- table(handBatAvg$bats)
numbats
```

#### Impact of Weight and Height on Fielding Percentage

```{r}
#Between Weight and Fielding Percentage
bodyFPweight <- ggplot(fieldingPercent, aes(x=FieldingPercentage, y=weight))+
  geom_point()+
  labs(title = "Comparison of Fielding Percentage and Players' Weight",
       x ="Fielding Percentage",
       y="Players' Weight")+
  theme_minimal(
  )

bodyFPweight

#Between Height and Fielding Percentage
bodyFPheight <- ggplot(fieldingPercent, aes(x=FieldingPercentage, y=height))+
  geom_point()+
  labs(title = "Comparison of Fielding Percentage and Players' Height",
       x ="Fielding Percentage",
       y="Players' Height")+
  theme_minimal(
  )

bodyFPheight

#Between Weight, Height, and Homeruns
bodyFPboth <- ggplot(fieldingPercent, aes(x=height, y=weight, color = FieldingPercentage, size = FieldingPercentage))+
  geom_point(alpha=0.7)+
  labs(title = "Comparison of Fielding Percentage and Players' Height and Weight",
       x ="Players' Height",
       y ="Players' Weight")+
  theme_minimal(
  )
bodyFPboth

#3D version of bodyFPboth
bodyFPboth3D <- plot_ly(x=fieldingPercent$height,y=fieldingPercent$weight,z=fieldingPercent$FieldingPercentage) |>
  layout(scene = list(
    xaxis = list(title = "Height"),
    yaxis = list(title = "Weight"),
    zaxis = list(title = "FieldingPercentage")))

bodyFPboth3D
```

## Communicate

#### Impact of Weight and Height on Homeruns

We have created insightful plots with GGplot and plot_ly to provide a visually compelling representation of our questions. Through these visualizations, we aim to find the relationship between weight, height, and the number of homeruns of the selected 50 baseball players during their careers.

We expected that taller and heavier hitters would record higher numbers of the sum of homeruns during their careers. However, our expectation was not met from the dataset.

It is evident from our analysis that the correlation between weight and height does not invariably dictate one's ability to achieve a high homerun count; however, it is noteworthy to acknowledge that there may be instances where these factors play a role in influencing such outcomes. When we increase our samples, there could be a relationship between weight, height, and the number of homeruns. Since there are only 50 samples, we might not be able to find a relationship.

#### Batting Average by Handedness

From the boxplot we can see that the batting average of left-handed hitter is the highest. The mean of the batting average of left-handed is 0.299. And the batting average of right-handed hitter is the second highest. The mean of the batting average of right-handed is 0.290. For the lowest batting average, both-handed is lowest. The mean of the batting average of both-handed is 0.286.

We expected that left-handed's batting average would be highest and both-handed's batting average would be lowest. And indeed, our expectations were correct.

There are a lot of left-handed hitters these days, but since there have been a lot more right-handed hitters, left-handed hitters may be able to target pitchers better because pitchers have relatively little experience with left-handed hitters. And left-handed hitters are more likely to hit infield than right-handed hitters because the first base is closer.

A both-handed hitter can play with his right hand when his opponent is a left-handed pitcher, and with his left hand when he is a right-handed pitcher, but that doesn't mean they are better than highly trained left-handed hitters, or highly trained right-handed hitters. In fact, as long as they don't use both hands well with real talent, they're likely to just be an ambiguous both-handed hitter.

#### Impact of Weight and Height on Fielding Percentage

We expected that shorter and heavier hitters would record lower fielding percentage during their careers. However, our expectation was not met from the dataset.

After creating plots to compare the effect of weight and height on fielding percentage, we can analyze the correlation to some degree. Based on these plots, there is no correlation.
